---
title: "Problem Set 7"
subtitle: "Due date: 30 October"
format: 
  html:
    self-contained: true
toc: true
editor: visual
execute: 
  echo: true
  warning: false
---

Please upload your completed assignment to the ELMs course site (under the assignments menu) and to your class Github repository. You need to upload both your Quarto document (with R code, as needed) and its rendered output (either as a PDF or HTML file).

**Total points: 20**

## Question 1

*Total points: 10*

I hypothesize that the older someone is, the more they like watching golf. I conduct a survey based on an SRS of adults in America. The independent variable is a four-category measure of age: 21-35, 36-50, 51-65, 66-plus. The dependent variable is a three-category measure of how much they enjoy watching golf: not at all, somewhat, very much.

### Part A

*Points: 4*

Make a cross-tab, by hand, of the relationship between these two variables based on the following (completely made up) data. Make sure that you present both raw counts and the appropriate percentages to examine the relationship.

```{r}
library(modelsummary)
library(tidyverse)

age_golf_df <- tibble::tibble(
  age = c("21-35", "51-65", "66-plus", "21-35", "36-50", "21-35", "36-50",
          "66-plus", "36-50", "51-65", "51-65", "21-35", "36-50", "21-35",
          "66-plus", "51-65", "36-50", "66-plus", "51-65", "21-35", "66-plus",
          "21-35", "36-50", "66-plus", "21-35", "36-50", "21-35", "66-plus",
          "51-65", "51-65", "66-plus", "21-35", "66-plus", "21-35", "51-65",
          "66-plus", "21-35", "36-50", "36-50", "66-plus"),
  preference = c("Somewhat", "Not at all", "Very much", "Not at all", "Somewhat",
                 "Not at all", "Not at all", "Somewhat", "Very much", "Very much",
                 "Somewhat", "Not at all", "Somewhat", "Very much", "Not at all",
                 "Not at all", "Somewhat", "Somewhat", "Very much", "Somewhat",
                 "Very much", "Somewhat", "Very much", "Somewhat", "Not at all",
                 "Very much", "Somewhat", "Somewhat", "Somewhat", "Very much",
                 "Not at all", "Not at all", "Very much", "Somewhat", "Very much",
                 "Very much", "Somewhat", "Somewhat", "Somewhat", "Not at all")
)

age_golf_df

datasummary_crosstab(preference ~ age, data = age_golf_df, statistic = 1 ~ 1 + N + Percent("row"))
```

### Part B

*Points: 2*

Do the data support my hypothesis? Be sure to explain the nature of the relationship (or lack thereof, if relevant).

**Answer**: There is weak support for the data. While there is slight positive relationship in older age rangers and "very much" preferencei in playing gold, the sample size is quite small, making it difficult to concretely determine the strength of the relationship.

### Part C

*Points: 2*

Compute (by hand) the chi-square statistic to test the null hypothesis of no relationship between these two variables. Be sure to show your work.

```{r}
obs_values <- count(age_golf_df, preference, age, name = "obs_n")
age_totals <- count(age_golf_df, age, name = "age_total")
preference_totals <- count(age_golf_df, preference, name = "preference_total")

obs_exp_counts <- preference_totals |> 
  expand_grid(age_totals) |> 
  relocate(age) |> 
  mutate(exp_n = (preference_total * age_total) / nrow(age_golf_df)) |>
  left_join(obs_values, by = c("age", "preference"))
obs_exp_counts

chi_sq <- obs_exp_counts |> 
  mutate(diff = obs_n - exp_n,
         diff_2 = diff^2,
         diff_2_standard = diff_2 / exp_n) |> 
  summarise(chi_sq = sum(diff_2_standard)) |> 
  pull()

print(chi_sq)
```

**Answer**: The chi-square is **6.09**.

### Part D

*Points: 2*

Using the chi-square statistic that you computed in question 1(c), can you reject the null hypothesis of no relationship between these two variables with 95% confidence? Why, or why not?

**Answer**: This chi-square value, with six degrees of freedom, produces a p-value of `r pchisq(chi_sq, df = 6, lower.tail = F)`, meaning the relationship observed is *not* statistically significant. Therefore, we cannot reject the null with 95% confidence. 

## Question 2

*Points: 10*

I hypothesize that people with higher levels of education tend to be more likely to pay higher levels of attention to politics. Use the `nes` dataset from the `poliscidols` package to test my hypothesis. Create a cross-tab examining the relationship between individuals' levels of education (`education`) and the time they spend paying attention to politics (`attention_to_politics`).

Do the data support my hypothesis? Be sure to explain the nature of the relationship (or lack thereof, if relevant). Report a chi-square statistic. What does the chi-square statistic tell you about whether you can reject the null hypothesis of no relationship between these two variables with 95% confidence? Why, or why not?

```{r}
nes_clean <- polisciols::nes %>%
   mutate(across(c(education, attention_to_politics), as.character)) %>%
   filter(!education %in% c("Refused", "Don't know"),
          !attention_to_politics %in% c("Refused", "Don't know"),
          !is.na(education),
          !is.na(attention_to_politics)) |> 
  mutate(education = factor(education, 
                            levels = c("Less than high school credential", 
                                       "High school credential", 
                                       "Some post-high school, no bachelor’s degree", 
                                       "Bachelor’s degree", 
                                       "Graduate degree")),
                            attention_to_politcs = factor(attention_to_politics,
                                       levels = c("Never",
                                                  "Some of the time",
                                                  "About half the time",
                                                  "Most of the time",
                                                  "Always"),
                                       ordered = TRUE))

datasummary_crosstab(attention_to_politics ~ education, data = nes_clean,
                     statistic = 1 ~ 1 + N + Percent("col"))

obs_values <- count(nes_clean, attention_to_politics, education, name = "obs_n")
  
attention_totals <- count(nes_clean, attention_to_politics, name = "attention_total")

education_totals <- count(nes_clean, education, name = "education_total")

obs_exp_counts <- attention_totals |> 
  expand_grid(education_totals) |> 
  relocate(education) |> 
  mutate(exp_n = (education_total * attention_total) / nrow(nes_clean)) |>
  left_join(obs_values, by = c("education", "attention_to_politics"))

q2_chi_sq <- obs_exp_counts |> 
  mutate(diff = obs_n - exp_n,
         diff_2 = diff^2,
         diff_2_standard = diff_2 / exp_n) |> 
  summarise(chi_sq = sum(diff_2_standard)) |> 
  pull()
q2_chi_sq

df <- (5 - 1) * (5 - 1)

pchisq(q2_chi_sq, df = df, lower.tail = F)

```
**Answer**:
The data suggests some support for the hypothesis. Levels of attention in "Most of the time" and "Always" tend to steadily increase from "less than high school credential" to "Graduate degree", suggesting that attention level slightly increases alongside education. 

The chi square value of 330.7615 produces a p-value of `r pchisq(330.7615, df = 16, lower.tail = F)`, indicating that the findings are very unlikely to be withinthe  normal distribution of the null and are thus statistically significant. Therefore, we can reject the null hypothesis. 
