---
title: "Problem Set 5"
subtitle: "Due date: 23 October"
format: 
  html:
    self-contained: true
toc: true
editor: visual
execute: 
  echo: true
---

Please upload your completed assignment to the ELMs course site (under the assignments menu) and to your class Github repository. You need to upload both your Quarto document (with R code, as needed) and its rendered output (either as a PDF or HTML file).

**Total points: 25**

```{r}
library(scales)
```

## Question 1

*Total points: 6*

Use the data in the table below to answer the following questions.

```{r}
#| echo: false

tibble::tibble(
  `Age group` = c("18-24", "25-30", "31 and up", "TOTAL"),
  `Non-voters` = c(70, 40, 220, 330),
  Voters = c(50, 50, 570, 670),
  Total = `Non-voters` + Voters
) |> 
  knitr::kable(caption = "Voting by Age in 2000")
```

### Part A

*Points: 2*

::: callout-note
2/2
:::

What is the probability of being 25-30 or a non-voter?

```{r}
((90/1000)+(330/1000))-(40/1000)
```

**Answer**: 0.38

### Part B

*Points: 4*

::: callout-note
1/4
:::

Assuming a normal distribution, report the 95% confidence intervals for the percentage of 18-to-24-year-olds who did not vote, and then the percentage of 25-to-30-year-olds who did not vote.

```{r}
p_hat_1824 <- (70/120)
p_hat_1824

se_1824 <- sqrt(((p_hat_1824 * (1 - p_hat_1824)) / 120))
se_1824

lower_1824 <- p_hat_1824 - 1.96 * se_1824
lower_1824

upper_1824 <- p_hat_1824 + 1.96 * se_1824
upper_1824
```

**Answer**: The confidence intervals for 18-to-24 year olds is **49.51%** and **67.15%**

```{r}
p_hat_2530 <- (40/90)
p_hat_2530

se_2530 <- sqrt(((p_hat_2530 * (1 - p_hat_2530)) / 90))
se_2530

lower_2530 <- p_hat_2530 - 1.96 * se_2530
lower_2530

upper_2530 <- p_hat_2530 + 1.96 * se_2530
upper_2530
```

**Answer**: The confidence intervals for 25-to-30-year-olds is **34.18%** and **54.71%**

::: callout-tip
The probability that an individual is 18-to-24 and a non-voter is not conditional. Therefore, you look at the proportion of the whole population who meet those criteria.

$$
\hat{p} = Pr(18\ to\ 24 \& NV) = \frac{70}{1000} = 0.07 
$$

And:

$$
n = 1000
$$

Therefore:

$$
SE(\hat{p}) = \sqrt{\frac{0.07 * 0.93}{1000}} = 0.008068457
$$

The 95% confidence interval can be found:

$$
\hat{p} \pm 1.96*SE(\hat{p})
$$

Therefore, the 95% confidence interval for the percentage of 18-to-24-year-olds who did not vote is `r percent(0.07 - 1.96 * 0.008068457, accuracy = 0.01)` to `r percent(0.07 + 1.96 * 0.008068457, accuracy = 0.01)`.

Using the same method, the 95% confidence intervals for the percentage of 25-to-30-year-olds who did not vote is `r percent(0.04 - 1.96*0.006196773, accuracy = 0.01)` to `r percent(0.04 + 1.96*0.006196773, accuracy = 0.01)`.
:::

## Question 2

*Total points:* *7*

Assume that the standard deviation for the population distribution of a state in which you want to conduct a poll is 200.

### Part A

*Points: 3*

::: callout-note
3/3
:::

Calculate the spread of the sampling distribution for each of the following sample sizes: 1, 4, 25, 100, 250, 1000, 5,000, and 10,000.

```{r}
(200/sqrt(1))
(200/sqrt(4))
(200/sqrt(25))
(200/sqrt(100))
(200/sqrt(250))
(200/sqrt(1000))
(200/sqrt(5000))
(200/sqrt(10000))
```

**Answer**:\
1 = 200\
4 = 100\
25 = 40\
100 = 20\
250 = 12.65\
1000 = 6.32\
5000 = 2.83\
10000 = 2

### Part B

*Points: 1*

::: callout-note
0.5/1
:::

Describe specifically how the variability of the sampling distribution changes as the sample size varies. Considering the expense of running a poll, which sample size do you think is most optimal if conducting the poll?\
\
**Answer**: 1000 offers a tight spread. Going past 1000 would only provide extremely diminished returns.

::: callout-note
Missing: Describe specifically how the variability of the sampling distribution changes as the sample size varies.
:::

### Part C

*Points: 3*

::: callout-note
2.5/3
:::

Display your results graphically (using R) with the sample size on the x-axis and the standard error (of the sampling distribution) on the y-axis.

```{r}
library(ggplot2)
library(tidyverse)

tibble(sample_size = seq(from = 200, to = 10000, by = 100)) |> 
  rowwise() |> 
  mutate(se = 200/sqrt(sample_size)) |> 
  ggplot(aes(x = sample_size, y = se)) + 
  geom_line() + 
  theme_minimal() + 
  labs(title = "Standard error according to sample size",
      x = "Number of observations in the sample",
      y = "Standard error")
```

::: callout-note
This is good! But the question asks you to display *your results.*
:::

## Question 3

*Points: 4*

::: callout-note
4/4
:::

Suppose you conduct a survey (to generate a sample mean of interest) and find that it has a margin of error of 4.5 with a sample size of 900 using a 95% confidence interval. What would the margin of error be for a 90% confidence interval?\
\
**Answer:\
**To start, we can plug in the given values in the following equation:\
`n = ( z2 * s2 ) / ( m2 )`

We find that we have the following:\
`900 = ( 1.962 * s2 ) / ( 4.52 )`\
\
We can solve for s^2^ and then calculate its square root.\
`( 900 * 4.52 ) / ( 1.962 ) = s2 )`\

```{r}
print(paste(
  sqrt(( 900 * 4.5^2 ) / ( 1.96^2 ))))

```

Now that we have s, we can calculate the m for a 90% confidence interval.\
`√n = ( √z2 * √s2 ) / ( √m2 )`\
\
To solve for the MOE, the equation becomes the following:\
`m = ( z * s ) / √n`

The z-score for a confidence interval of 90% is approximately 1.645. Given that, the equation to solve for the new m is as follows:\

```{r}
print(paste(
  (1.645 * 68.87755) / sqrt(900)))
```

The margin of error for a 90% confidence interval is approximately **3.78**.

## Question 4

*Points: 4*

::: callout-note
4/4
:::

Assume that, in State A, the mean income in the population is \$20,000 with a standard deviation of \$2,000. If you took an SRS of 900 individuals from that population, what is the probability that you would get a sample mean income of \$20,200 or greater? What would be the probability if the sample size was only 25?

::: callout-note
Assume a normal distribution for both questions.
:::

**Answer**:\
First, we have to calculate the standard error. We can use the SE for both survey sample sizes to determine their respective Z scores. This is done through the following equation:\
`Z = x̄ - µ / σ`\
Since these surveys are assumed to be distributed Normally, we can use these Z-scores to determine probability.\

```{r}
s_900 <- (2000/sqrt(900))

z_900 <- ((20200 - 20000)/s_900)

print(z_900)

s_25 <- (2000/sqrt(25))

z_25 <- ((20200 - 20000)/s_25)

print(z_25)

```

P(Z ≥ 3.0) = 0.00135\
The probability that you would get a sample mean income of \$20,200 or greater in a SRS of 900 is 0.00135.\
\
P(Z ≥ 0.5) = 0.30854\
The probability that you would get a sample mean income of \$20,200 or greater in a SRS of 25 is 0.30854.

## Question 5

*Points: 4*

::: callout-note
1/4
:::

Assume that a coin is fair. If I flip a coin 500 times, what is a 95% confidence interval for the range of the count of heads that I will get? What if I flip the coin 5,000 times? What about 50,000 times?

**Answer**:\
First, we need to find the standard error for these different counts using the following equation:\
`se(p̂) = √((p (1 - p)/n)`

Since these coins are assumed to be fair, the probability of either outcome is 0.5. Therefore, p = 0.5. Once these standard errors are accounted for, the following equation can be used to determine the range of the count of heads within a 95% confidence interval:\
`p̂ ± 1.96 * se(p̂)`\

```{r}
se_500 <- sqrt((0.5 * 0.5)/500)

upper_500 <- (500 + 1.96*se_500)
lower_500 <- (500 - 1.96*se_500)

print(paste("The range of the count of heads within a 95% confidence interval for 500 coin flips is between", upper_500, "and", lower_500))

se_5k <- sqrt((0.5 * 0.5)/5000)

upper_5k <- (5000 + 1.96*se_5k)
lower_5k <- (5000 - 1.96*se_5k)

print(paste("The range of the count of heads within a 95% confidence interval for 5000 coin flips is between", upper_5k, "and", lower_5k))

se_50k <- sqrt((0.5 * 0.5)/50000)

upper_50k <- (50000 + 1.96*se_50k)
lower_50k <- (50000 - 1.96*se_50k)

print(paste("The range of the count of heads within a 95% confidence interval for 50000 coin flips is between", upper_50k, "and", lower_50k))
```

::: callout-note
The standard deviation of a binomial distribution:

$$
\sigma_X = \sqrt{np(1-p)}
$$
:::
