---
title: "Problem Set 4"
subtitle: "Due date: 9 October"
format: 
  html:
    self-contained: true
toc: true
editor: visual
execute: 
  echo: true
---

Please upload your completed assignment to the ELMs course site (under the assignments menu) and to your class Github repository. You need to upload both your Quarto document (with R code, as needed) and its rendered output (either as a PDF or HTML file).

**Total points: 30**

## Question 1

*Points: 5*

Suppose I am playing poker with a regular deck of 52 cards. I have three cards in my hand, one of which is a king. What is the probability that the next two cards I draw will both be kings?

**Answer**:

```{r}
(3/49)*(2/48)
```

0.0026 or .26% chance that the next two cards with both be kings.

## Question 2

*Points: 5*

A survey of 1,500 randomly selected individuals reveals that 46% of them have a college degree and that 31% of them have a gun in their house. If you were to assume that college graduation and gun ownership are independent of one another, what is the probability that an individual is both a college graduate and does not have a gun in their house?\

**Answer: \
**P(A & B) = P(A)\*P(B)

```{r}
(.46)*(1.00-.31)
```

P = 0.317 *or* 31.7% chance that an individual a college graduate that does not own a firearm.

## Question 3

*Points: 5*

I roll a fair six-sided dice and I get a number greater than three. What is the probability that the next two dice rolls will also generate numbers greater than three?\
\
**Answer**:\
P(A & B) = P(A)\*P(B)

```{r}
(1/2)*(1/2)
```

P = 0.25 *or* 25% chance that the next two dice rolls will both be greater than three.

## Question 4

*Points: 10*

Assume that in the United States, 15% of people are smokers and, of smokers, 60% smoke at least one pack a day. Among those who smoke at least a pack a day, 60% will develop lung cancer at some point in their lives while, in the rest of the population (which includes smokers who smoke less than a pack a day and non-smokers), only 10% will. What is the probability that a person who develops lung cancer was a smoker who smoked at least a pack a day?\
\
**Answer**:\
First we need to find the total proportion of smokers, in this case *casual/non-smoker* and *habitual*, who develop lung cancer in relation to the general population. \
P(A) is the probability that a habitual or one pack-a-day smokers will develop lung cancer \
P(B) is the probability that a casual or less than one pack-a-day smokers will develop lung cancer.\
P(C) is the probability that someone from the rest of the population will develop lung cancer.\
For the sake of this question, we can simplify the general population to 100.\
\
To calculate the probability of a habitual smoker developing lung cancer:\
`P(A) = (.15)*(.60)*(.60)`

```{r}
(.15)*(.60)*(.60)
```

P(B) = .054 *or* 5.4% chance that someone is a habitual smoker who will develop lung cancer. With the proposed population, this totals out to 54.\
\
To calculate the probability of a casual smoker developing lung cancer:\
`P(B) = (.15)*(.40)*(.10)`

```{r}
(.15)*(.40)*(.10)
```

P(B) = .006 or 0.6% chance that someone is a casual smoker who will develop lung cancer. With the proposed population, this totals out to 6.\
\
To calculate the probability that the rest of the population will develop lung cancer:\
`P(C) = (.85)*(.10)`

```{r}
(.85)*(.10)
```

P(RLC) = 0.085 *or* 8.5% chance that the remaining population will develop lung cancer. With the proposed population, this totals out to 85.\
\
The following formula will determine the probability someone who develops lung cancer is a habitual smoker:\
`P(D) = P(A)/(P(A)+P(B)+P(C))`\

```{r}
(.054)/(.054+.006+.085)
```

P(D) = 0.372 *or* 37.2% chance that someone who develops cancer lung cancer is a habitual smokers.

## Question 5

*Points: 5*

Assume that in the world, 60% of countries are democracies and 80% of countries have ratified the Rome Statute for the International Criminal Court. Assume further that 90% of democracies have ratified the Rome Accords. What is the probability that a country is either a democracy or has ratified the Rome Accords?\

**Answer**:\
P(A) represents the probability that a country is a democracy\
P(B) represents the probability that a country has ratified the Rome Statute\
P(A and B) represents the probability that a democracy has ratified the Rome Accords\
Given this information, we can calculate P(A or B) by accounting for the union:\
`P(A or B) = P(A) + P(B) - P(A and B)`\

```{r}
(.60) + (.80) - (.60*.90)
```

P(A or B) = 0.86 *or* there is a 86% chance a country is either a democracy or has ratified the Rome Accords.
